---
title: "Chapter 8: Group comparisons and hierarchical modeling"
author: "Jesse Mu"
date: "November 17, 2016"
output:
  html_document:
    highlight: pygments
    toc: yes
    toc_float: yes
---

<!-- Setup -->

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { 
      equationNumbers: {
            autoNumber: "all"
      }
  }
});
</script>

```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(fig.align = 'center', message = FALSE)
library(knitr)
library(ggplot2)
library(cowplot)
library(reshape)
```


# Hierarchical modeling of means and variances

The previous model assumed common variance within groups $\sigma^2$. This is 
actually fairly common, perhaps less because of empirical justification for 
assuming common within-group variance than lack of interest in the variance of 
the groups. But of course, the inaccuracy of this assumption could result in 
errors in analysis. It's fairly straightforward to simply add another 
hierarchical layer for the variance too, and jointly estimate the group-specific
means and variances, as well as the common mean and variance parameters.

To implement this, we let $\theta_j$ depend on $\boldsymbol{y}_j$ and (new!) a
group $j$-specific $\sigma_j^2$, so our full conditional distribution is

$$
\theta_j \mid \boldsymbol{y}_j, \sigma_j^2 \sim \mathcal{N}\left( \frac{n_j\bar{y}_j / \sigma_j^2 + 1 / \tau^2}{n_j / \sigma_j^2 + 1 / \tau^2}, \left[n_j / \sigma_j^2 + 1 / \tau^2 \right]^{-1} \right)
$$

similarly, above we had a rather special case of the full conditional of
$\sigma^2$. Since there was one common $\sigma^2$, the posterior was based on a 
combination of the prior precision and the pooled sample variance. Now let's 
assume that we have a separate $\sigma_j^2$ for each group:

$$\sigma_1^2, \dots, \sigma_m^2 \sim  \text{i.i.d.} \; \mathcal{N}(\nu_0 / 2, \sigma_0^2 \nu_0 / 2).$$

Now that we're including individual $\sigma_j^2$, re-deriving the full
conditional for $\sigma_j^2$ results in full conditional distributions that look
just like the one-parameter case for variance (and the corresponding $\theta_j$ conditionals for the means):

$$
\sigma_j^2 \mid \boldsymbol{y}_j, \theta_j \sim \text{inverse-gamma}\left([\nu_0 - n_j] / 2, \left[\nu_0 \sigma_0^2 + \sum_{i = 1}^{n_j}(y_{i, j} - \theta_j)^2 \right] / 2 \right)
$$

Now in the same way that we learn the group-specific parameters $\mu, \tau^2$ 
using prior distributions with hyperparameters $\eta_0, \tau_0^2, \mu_0, 
\gamma_0^2$, we should learn $\nu_0, \sigma_0^2$ with prior distributions.

Since we are modeling variances, we let $\sigma_0^2 \sim \text{gamma}(a, b)$ such that the posterior is

$$
\sigma_0^2 \mid \sigma_1^2, \dots, \sigma_m^2, \nu_0 \sim \text{gamma}\left(a + \frac{1}{2}m \nu_0, b + \frac{1}{2}\sum_{j = 1}^m (1 / \sigma_j^2)\right)
$$

Lastly, there is no simple conjugate prior for $\nu_0$ unless we restrict
$\nu_0$ to be a whole number. If we let $\nu_0 \sim \text{geometric}(\alpha)$,
then (as the discrete analog of the exponential distribution) $p(\nu_0) \sim
\text{exp}(-\alpha \nu_0)$ and the full conditional distribution of $\nu_0$ can be shown to be

$$
p(\nu_0 \mid \sigma_0^2, \dots, \sigma_m^2) \propto \left( \frac{(\sigma_0^2 \nu_0 / 2)^{\nu_0 / 2}}{\Gamma(\nu_0 / 2)} \right)^m \left( \prod_{j = 1}^m \frac{1}{\sigma_j^2} \right)^{\nu_0 / 2 - 1} \times \text{exp} \left( - \nu_0 \left(\alpha + \frac{1}{2} \sigma_0^2 \sum_{j = 1}^m \frac{1}{\sigma_j^2} \right) \right)
$$

## Analysis of math score data

For time I haven't reproduced the analysis here are there aren't too many
conclusions drawn - the point is just that you can let the variance vary across
groups.

# Exercises

## 8.1

### a

I expect $\text{Var}(y_{i, j} \mid \mu, \tau^2)$ to be bigger since it includes
both within- and between-group sampling variability.

### b

I think $\text{Cov}(y_{i_1, j}, y_{i_2, j} \mid \theta_j, \sigma^2)$ is zero
because, according to exchangeability, our $y_{i, j}$ are conditionally i.i.d.
when $\theta_j, \sigma^2$ is known.

On the other hand, given our model, it seems like knowing about another $y_{i_1,
j}$ *does* provide more information about $y_{i_2, j}$, and I expect them to
covary positively with each other. Specifically, $y_{i_1, j}$ seems to give more
information about what the mean $\theta_j$ is, and we expect values from the
same $\theta_j$ to be closer together (due to decreased variability). I can't
come up with a more formal mathematical justification, though.

### c

\begin{align}
\text{Var}(y_{i, j} \mid \theta_j, \sigma^2) &= \sigma^2 & \text{By def.}\\
\text{Var}(\bar{y}_{\cdot, j} \mid \theta_j, \sigma^2) &= \sigma^2 / n_j & \text{Samp. dist. mean} \\
\text{Cov}(y_{i_1, j}, y_{i_2, j} \mid \theta_j, \sigma^2) &= \mathbb{E}(y_{i_1, j}y_{i_2, j}) - \mathbb{E}(y_{i_1, j})\mathbb{E}(y_{i_2, j}) \\
&= \mathbb{E}(y_{i_1, j})\mathbb{E}(y_{i_2, j}) - \mathbb{E}(y_{i_1, j})\mathbb{E}(y_{i_2, j}) & \text{i.i.d.} \\
&= 0 \\
& \\
\text{Var}(y_{i, j} \mid \mu, \tau^2) &= \text{Var}(\mathbb{E}(y_{i, j} \mid \theta_j, \sigma^2) \mid \mu, \tau^2) + \mathbb{E}(\text{Var}(y_{i, j} \mid \theta_j, \sigma^2) \mid \mu, \tau^2) & \text{Law total var.} \\
&= \text{Var}(\theta_j \mid \mu, \tau^2) + \mathbb{E}(\sigma^2 \mid \mu, \tau^2) \\
&= \tau^2 + \sigma^2 \\

\text{Var}(\bar{y}_{\cdot, j} \mid \mu, \tau^2) &= \text{Var}(\mathbb{E}(\bar{y}_{\cdot, j} \mid \theta_j, \sigma^2) \mid \mu, \tau^2) + \mathbb{E}(\text{Var}(\bar{y}_{\cdot, j} \mid \theta_j, \sigma^2) \mid \mu, \tau^2) & \text{Law total var.} \\
&= \text{Var}(\theta_j \mid \mu, \tau^2) + \mathbb{E}(\sigma^2 / n_j \mid \mu, \tau^2) \\
&= \tau^2 + (\sigma^2 / n_j) \\

\text{Cov}(y_{i_1, j}, y_{i_2, j} \mid \mu, \tau^2) &= \text{E}(\text{Cov}(y_{i_1, j}, y_{i_2, j} \mid \theta_j, \sigma^2) \mid \mu, \tau^2) + \text{Cov}(\mathbb{E}(y_{i_1, j} \mid \theta_j, \sigma^2), \mathbb{E}(y_{i_2, j} \mid \theta_j, \sigma^2)) & \text{Law total covar.} \\
&= \text{E}(0 \mid \mu, \tau^2) + \text{Cov}(\mathbb{E}(y_{i_1, j} \mid \theta_j, \sigma^2), \mathbb{E}(y_{i_2, j} \mid \theta_j, \sigma^2)) & \text{i.i.d.} \\
&= \text{Cov}(\theta_j, \theta_j) \\
&= \text{Var}(\theta_j) \\
&= \tau^2
\end{align}

These values indeed align with the intuitions above. The values for the
variances and covariances with $\theta_j$ unknown are simply those for
$\theta_j$ known plus $\tau^2$, the between-group sampling variability.

### d

For convenience let $\mathcal{D} = \{\boldsymbol{y}_1, \dots,
\boldsymbol{y}_m\}$ and $\boldsymbol{\theta} = \{ \theta_1, \dots, \theta_m \}$.

Also, if we treat the model as a Bayes' net, we can use factorization to quickly
extract the conditional independencies: $P(X_1, \dots, X_n) = \prod_{i = 1}^n P(X_i \mid \text{Pa}(X_i))$ where $\text{Pa}(X)$ are the parents of $X$.

\begin{align}
p(\mu \mid \mathcal{D}, \boldsymbol{\theta}, \sigma^2, \tau^2) &= \frac{p(\mu, \mathcal{D}, \boldsymbol{\theta}, \sigma^2, \tau^2)}{\int p(\mu, \mathcal{D}, \boldsymbol{\theta}, \sigma^2, \tau^2) \; d\mu} \\
&= \frac{p(\mu) p(\tau^2) p(\sigma^2) p(\mathcal{D} \mid \boldsymbol{\theta}, \sigma^2) p(\boldsymbol{\theta} \mid \mu, \tau^2) } {\int p(\mu) p(\tau^2) p(\sigma^2) p(\mathcal{D} \mid \boldsymbol{\theta}, \sigma^2) p(\boldsymbol{\theta} \mid \mu, \tau^2)\; d\mu } & \text{Factorization} \\
&= \frac{p(\mu) p(\tau^2) p(\sigma^2) p(\mathcal{D} \mid \boldsymbol{\theta}, \sigma^2) p(\boldsymbol{\theta} \mid \mu, \tau^2) } { p(\tau^2) p(\sigma^2) p(\mathcal{D} \mid \boldsymbol{\theta}, \sigma^2) \int p(\mu) p(\boldsymbol{\theta} \mid \mu, \tau^2)\; d\mu } & \text{Constants outside} \\
&= \frac{p(\mu) p(\boldsymbol{\theta} \mid \mu, \tau^2) } { \int p(\mu) p(\boldsymbol{\theta} \mid \mu, \tau^2)\; d\mu } & \\
&= p(\mu \mid \boldsymbol{\theta}, \tau^2) & \text{Bayes' rule}
\end{align}

This means that $\mu$ does not depend on the data (or $\sigma^2$) once
$\theta_1, \dots, \theta_m$ are known; another example of conditional
independence induced by the Bayes network.

## 8.3

### a

```{r}
# Load data
library(dplyr)
library(tidyr)
schools.list = lapply(1:8, function(i) {
  f = paste("school",i,".dat",sep="")
  w = read.table(f)
  
  data.frame(
    school = i,
    hours = w[, 1] %>% as.numeric
  )
})
schools.raw = do.call(rbind, schools.list)

Y = schools.raw


# Prior
mu0 = 7
g20 = 5
t20 = 10
eta0 = 2
s20 = 15
nu0 = 2

# Number of schools. Y[, 1] are school ids
m = length(unique(Y[, 1]))

# Starting values - use sample mean and variance
n = sv = ybar = rep(NA, m)
for (j in 1:m) {
  Y_j = Y[Y[, 1] == j, 2]
  ybar[j] = mean(Y_j)
  sv[j] = var(Y_j)
  n[j] = length(Y_j)
}

# Let initial theta estimates be the sample means
# Similarly, let initial values of sigma2, mu, and tau2 be "sample mean and
# variance"
theta = ybar
sigma2 = mean(sv)
mu = mean(theta)
tau2 = var(theta)

# MCMC
S = 1500
THETA = matrix(nrow = S, ncol = m)
# Storing sigma, mu, theta together
SMT = matrix(nrow = S, ncol = 3)
colnames(SMT) = c('sigma2', 'mu', 'tau2')

for (s in 1:S) {
  # Sample thetas
  for (j in 1:m) {
    vtheta = 1 / (n[j] / sigma2 + 1 / tau2)
    etheta = vtheta * (ybar[j] * n[j] / sigma2 + mu / tau2)
    theta[j] = rnorm(1, etheta, sqrt(vtheta))
  }
  
  # Sample sigma2
  nun = nu0 + sum(n) # TODO: Could cache this
  ss = nu0 * s20
  # Pool variance
  for (j in 1:m) {
    ss = ss + sum((Y[Y[, 1] == j, 2] - theta[j])^2)
  }
  sigma2 = 1 / rgamma(1, nun / 2, ss / 2)
  
  # Sample mu
  vmu = 1 / (m / tau2 + 1 /g20)
  emu = vmu * (m * mean(theta) / tau2 + mu0 / g20)
  mu = rnorm(1, emu, sqrt(vmu))
  
  # Sample tau2
  etam = eta0 + m
  ss = eta0 * t20 + sum((theta - mu)^2)
  tau2 = 1 / rgamma(1, etam / 2, ss / 2)
  
  # Store params
  THETA[s, ] = theta
  SMT[s, ] = c(sigma2, mu, tau2)
}
```

Assess convergence with diagnostic boxplots:

```{r echo=FALSE}
smt.df = data.frame(SMT)
colnames(smt.df) = c('sigma2', 'mu', 'tau2')
smt.df$s = 1:S

cut_size = 10

smt.df = smt.df %>%
  tbl_df %>%
  mutate(scut = cut(s, breaks = cut_size)) %>%
  gather('variable', 'value', sigma2:tau2)

ggplot(smt.df, aes(x = scut, y = value)) +
  facet_wrap(~ variable, scales = 'free_y') +
  geom_boxplot() +
  theme(axis.text.x = element_blank()) +
  xlab('Samples')
```

Evaluate effective sample size:

```{r}
# Tweak number of samples until all of the below are above 1000
library(coda)
effectiveSize(SMT[, 1])
effectiveSize(SMT[, 2])
effectiveSize(SMT[, 3])
```

### b

Posterior means and confidence intervals

```{r}
t(apply(SMT, MARGIN = 2, FUN = quantile, probs = c(0.025, 0.5, 0.975)))
```

Comparing posterior to prior:
```{r}
# For dinvgamma
library(MCMCpack)
sigma2_prior = data.frame(
  value = seq(10, 22.5, by = 0.1),
  density = dinvgamma(seq(10, 22.5, by = 0.1), nu0 / 2, nu0 * s20 / 2),
  variable = 'sigma2'
)
tau2_prior = data.frame(
  value = seq(0, 30, by = 0.1),
  density = dinvgamma(seq(0, 30, by = 0.1), eta0 / 2, eta0 * t20 / 2),
  variable = 'tau2'
)
mu_prior = data.frame(
  value = seq(0, 12, by = 0.1),
  density = dnorm(seq(0, 12, by = 0.1), mu0, sqrt(g20)),
  variable = 'mu'
)
priors = rbind(sigma2_prior, tau2_prior, mu_prior)
priors$dist = 'prior'
smt.df$dist = 'posterior'

ggplot(priors, aes(x = value, y = density, color = dist)) +
  geom_line() +
  geom_density(data = smt.df, mapping = aes(x = value, y = ..density..)) +
  facet_wrap(~ variable, scales = 'free')
```

Our prior estimates for $\mu$ and $\tau^2$ were fairly estimate, but our 
estimate for $\sigma^2$ was very far off. After this analysis, we have estimates
for $\mu$, the average amount of hours of schoolwork spent at a typical school, 
$\tau^2$, the variability between schools in the average hours of schoolwork, 
and $\sigma^2$, the variability among students' hours in each school.

### c
```{r}
t20_prior = (1 / rgamma(1e6, eta0 / 2, eta0 * t20 / 2))
s20_prior = (1 / rgamma(1e6, nu0 / 2, nu0 * s20 / 2))

R_prior = data.frame(
  value = (t20_prior) / (t20_prior + s20_prior),
  dist = 'prior'
)
R_post = data.frame(
  value = SMT[, 'tau2'] / (SMT[, 'tau2'] + SMT[, 'sigma2']),
  dist = 'posterior'
)

ggplot(R_prior, aes(x = value, y = ..density.., color = dist)) +
  geom_density(data = R_prior) +
  geom_density(data = R_post)

mean(R_post$value)
```

$R$ measures how much of the total variance in our data is between-group. Our
prior didn't contain much information about this quantity, but after inference,
we expect that around 25\% of our variance comes from between group variance
($\tau^2$).

### d

```{r}
theta7_lt_6 = THETA[, 7] < THETA[, 6]
mean(theta7_lt_6)

theta7_smallest = (THETA[, 7] < THETA[, -7]) %>%
  apply(MARGIN = 1, FUN = all)

mean(theta7_smallest)
```

### e


```{r}
relationship = data.frame(
  sample_average = ybar,
  post_exp = colMeans(THETA),
  school = 1:length(ybar)
)
ggplot(relationship, aes(x = sample_average, y = post_exp, label = school)) +
  geom_text() +
  geom_abline(slope = 1, intercept = 0) +
  geom_hline(yintercept = mean(schools.raw[, 'hours']), lty = 2) +
  annotate('text', x = 10, y = 7.9, label = paste0("Pooled sample mean ", round(mean(schools.raw[, 'hours']), 2))) +
  geom_hline(yintercept = mean(SMT[, 'mu']), color = 'red') +
  annotate('text', x = 10, y = 7.4, label = paste0("Posterior exp. mu ", round(mean(SMT[, 'mu']), 2)), color = 'red')
```

There is a quite tight correspondence between the sample average and the
posterior expectation, although mild shrinkage can be observed with schools with
very high and low sample averages being pulled towards the mean.